The Weather Events Analysis
========================================================

This project try to answer two questions:
* Across the United States, which types of events (as indicated in the `EVTYPE` variable) are most harmful with respect to population health?
* Across the United States, which types of events have the greatest economic consequences?

The other detail about the project can be found on Coursera website.

#  Data Processing 
The raw data can't be read into R completly, the `REMARK` variable is too big to cotains in one column, so, I 
just read a part of row. I load just 7 variable which is `EVTYPE`,`FATALITIES`,`INJURIES`,`PROPDMG`,`PROPDMGEXP`,`CROPDMG` and `CROPDMGEXP`  
```{r,cache=TRUE}
StormData=read.csv("repdata-data-StormData.csv",sep=",",header=TRUE,stringsAsFactors=FALSE,nrows=547362,colClasses=
                     c(rep("NULL",7),"character",rep("NULL",14),rep("numeric",2),rep(c("numeric","character"),2),
                       rep("NULL",9)))
head(StormData)
```
Now, I will show you how I clean the data to be ready for data analysis, fist of all, I transfer the following three variables to capital letter.
```{r,cache = TRUE}

StormData$EVTYPE=toupper(StormData$EVTYPE)
StormData$PROPDMGEXP=toupper(StormData$PROPDMGEXP)
StormData$CROPDMGEXP=toupper(StormData$CROPDMGEXP)
```
The reduce data is named `StormDataV1` that remove the meanningless symbol in variabe `PROPDMGEXP` and `CROPDMGEXP` like ``, `+`, `-` and `?`. 
```{r,cache = TRUE}

index=!(StormData$CROPDMGEXP=="?" | StormData$CROPDMGEXP=="" | 
          StormData$PROPDMGEXP=="?" | StormData$PROPDMGEXP=="" | 
          StormData$PROPDMGEXP=="-" | StormData$PROPDMGEXP=="+")
StormDataV1=StormData[index,]
```
Change variabe `PROPDMGEXP` and `CROPDMGEXP` to numeric class which unit is USD.
```{r,cache = TRUE}
StormDataV1$PROPDMGEXP=ifelse(StormDataV1$PROPDMGEXP=="K",1000,
                              ifelse(StormDataV1$PROPDMGEXP=="M",1000000,
                                     ifelse(StormDataV1$PROPDMGEXP=="B",10^9,
                                            ifelse(StormDataV1$PROPDMGEXP==0,1,
                                                   ifelse(StormDataV1$PROPDMGEXP==3,10,10^5)))))
StormDataV1$CROPDMGEXP=ifelse(StormDataV1$PROPDMGEXP=="K",1000,
                              ifelse(StormDataV1$PROPDMGEXP=="M",1000000,
                                     ifelse(StormDataV1$PROPDMGEXP=="B",10^9,1)))     
```
There are numbers of different types of weather events, so, what I do next is to group it to 8 groups. The detail shows at the following code.
```{r,cache = TRUE}
tornado.index=grep("TORNADO|GUSTNADO|WATERSPOUT",StormDataV1$EVTYPE,ignore.case=TRUE)
tonado.class=unique((StormDataV1$EVTYPE)[tornado.index])
storm.index=grep("STORM|TYPHOON|HURRICANE|LIGHT",(StormDataV1$EVTYPE)[-tornado.index],ignore.case=TRUE)
storm.class=unique(((StormDataV1$EVTYPE)[-tornado.index])[storm.index])
wind.index=grep("WIND",((StormDataV1$EVTYPE)[-tornado.index])[-storm.index],ignore.case=TRUE)
wind.class=unique((((StormDataV1$EVTYPE)[-tornado.index])[-storm.index])[wind.index])
flood.index=grep("FLOOD|RAIN|FLD",(((StormDataV1$EVTYPE)[-tornado.index])[-storm.index])[-wind.index],ignore.case=TRUE)
flood.class=unique(((((StormDataV1$EVTYPE)[-tornado.index])[-storm.index])[-wind.index])[flood.index])
hail.index=grep("HAIL",((((StormDataV1$EVTYPE)[-tornado.index])[-storm.index])[-wind.index])[-flood.index],ignore.case=TRUE)
hail.class=unique((((((StormDataV1$EVTYPE)[-tornado.index])[-storm.index])[-wind.index])[-flood.index])[hail.index])
heat.index=grep("HEAT|FIRE|DRY|DROUGHT",(((((
  StormDataV1$EVTYPE)[-tornado.index])[-storm.index])[-wind.index])[-flood.index])[-hail.index],ignore.case=TRUE)
heat.class=unique(((((((StormDataV1$EVTYPE)[-tornado.index])[-storm.index])[-wind.index])[-flood.index])[-hail.index])
                  [heat.index])
snow.index=grep("SNOW|BLIZZARD|COLD|FREEZE|ICE|ICY",((((((
  StormDataV1$EVTYPE)[-tornado.index])[-storm.index])[-wind.index])[-flood.index])[-hail.index])[-heat.index],ignore.case=TRUE)
snow.class=unique((((((((StormDataV1$EVTYPE)
                        [-tornado.index])[-storm.index])[-wind.index])[-flood.index])[-hail.index])[-heat.index])
                  [snow.index])
other.class=unique((((((((StormDataV1$EVTYPE)
                         [-tornado.index])[-storm.index])[-wind.index])[-flood.index])[-hail.index])[-heat.index])
                   [-snow.index])

```
The clean data set `CleanData` has five variable, `EVGROUP` is available at previous step, `PropDmg` and `CropDmg` are simply `PROPDMG` times `PROPDMGEXP` and `CROPDMG` times `CROPDMGEXP`, `FATALITIES` and `INJURIES` are same as `StormDataV1` 
```{r,cache = TRUE}
EVGROUP=ifelse(StormDataV1$EVTYPE%in%tonado.class,"TORNADO",
                 ifelse(StormDataV1$EVTYPE%in%storm.class,"STORM",
                        ifelse(StormDataV1$EVTYPE%in%wind.class,"WIND",
                               ifelse(StormDataV1$EVTYPE%in%flood.class,"FLOOD",
                                      ifelse(StormDataV1$EVTYPE%in%hail.class,"HAIL",
                                      ifelse(StormDataV1$EVTYPE%in%heat.class,"HEAT",
                                             ifelse(StormDataV1$EVTYPE%in%snow.class,"SNOW","OTHER")))))))
FATALITIES=StormDataV1$FATALITIES
INJURIES=StormDataV1$INJURIES
PropDmg=StormDataV1$PROPDMG*StormDataV1$PROPDMGEXP
CropDmg=StormDataV1$CROPDMG*StormDataV1$CROPDMGEXP
CleanData=data.frame(EVGROUP,FATALITIES,INJURIES,PropDmg,CropDmg)
```
```{r,eche = TRUE}
head(CleanData)
```

# Method
The function `fatal.rate` and `injury.rate`  indicate the influence of population health. The rate is defined as **the percentage of the specific weather event causes  at least one died or injury**. 
```{r,cache = TRUE}

fatal.rate=function(x)
{
  sub.data=CleanData[CleanData[,1]==x,]
  sum(sub.data[,2]>0)/length(sub.data[,2])
}
injury.rate=function(x)
{
  sub.data=CleanData[CleanData[,1]==x,]
  sum(sub.data[,3]>0)/length(sub.data[,3])
}
```
The function `prop.dmg`, `crop.dmg` and `total.dmg`  indicate **the 
average of property damage, crop damage and total damage for specific weather event**.
```{r,cache = TRUE}

prop.dmg=function(x)
{
  mean(CleanData[CleanData[,1]==x,4])
}
crop.dmg=function(x)
{
  mean(CleanData[CleanData[,1]==x,5])
}
total.dmg=function(x)
{
  mean(CleanData[CleanData[,1]==x,5]+CleanData[CleanData[,1]==x,4])
}
```

# Results
We can get the numeric result in the following two tables. 
```{r}

FatalRate=apply(cbind(c("TORNADO","STORM","WIND","FLOOD","HAIL","HEAT","SNOW","OTHER")),1,fatal.rate)
InjuryRate=apply(cbind(c("TORNADO","STORM","WIND","FLOOD","HAIL","HEAT","SNOW","OTHER")),1,injury.rate)
health.table=as.data.frame(rbind(FatalRate,InjuryRate))
colnames(health.table)=c("TORNADO","STORM","WIND","FLOOD","HAIL","HEAT","SNOW","OTHER")
health.table
```
```{r}

PropertyDamage=apply(cbind(c("TORNADO","STORM","WIND","FLOOD","HAIL","HEAT","SNOW","OTHER")),1,prop.dmg)
CropDamage=apply(cbind(c("TORNADO","STORM","WIND","FLOOD","HAIL","HEAT","SNOW","OTHER")),1,crop.dmg)
TotalDamage=apply(cbind(c("TORNADO","STORM","WIND","FLOOD","HAIL","HEAT","SNOW","OTHER")),1,total.dmg)
damage.table=as.data.frame(rbind(PropertyDamage,CropDamage,TotalDamage))
colnames(damage.table)=c("TORNADO","STORM","WIND","FLOOD","HAIL","HEAT","SNOW","OTHER")
damage.table
```
The following figure combines the information of population health and economic consequences, **the size of text represent the degree of total damage**.
```{r}

text.size=as.numeric(3-(22392249-damage.table[3,])/8956780)
plot(FatalRate,InjuryRate,type="n")
text(FatalRate,InjuryRate,c("TORNADO","STORM","WIND","FLOOD","HAIL","HEAT","SNOW","OTHER"),cex=text.size,col=2)
text(0.01,0.15,"Bigger size, Higher total damage")
text(0.01,0.125,"Smaller size, Lower total damage")
```

Finally, we answer the original two questions
* **TORNADO is most harmful with respect to population health because the figure shows high fatal rate and high injury rate.**
* **STORM have the greatest economic consequences because the text size is biggest.**